---
title: "Main publication figures for Nawaz et al"
author: "urwahnawaz"
date: "2024-10-08"
output:
  html_notebook: default
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
	echo = TRUE,
	warning = FALSE,
	message = FALSE
)
```

# Introduction 

This notebook contains code used to generate main figures in [Nawaz et al (2025)](https://www.biorxiv.org/content/10.1101/2025.06.12.659426v1). The code used to generate the mRNA-protein correlation data used for these figures can be found in  `analysis/calculate_correlations.Rmd`, or in the Analysis option on the option bar in this notebook. Figures which utilise the raw protein or RNA expression matrices can be found in the relevant publicaitons. 


## Import libraries and set-up

```{r}
library(ggplot2)
library(magrittr)
library(tidyverse)
library(readxl)
library(dplyr)
library(RWDataPlyr)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(EnsDb.Hsapiens.v86)
library(svglite)
library(ggiraph)
library(reshape2)
library(ggvenn)
library(VennDiagram)
library(ggtext)
library(colorspace)
library(ragg)
library(msigdbr)
library(matrixStats)
library(ggpubr)
```

```{r functions}
library(readxl)    
read_excel_allsheets <- function(filename, tibble = FALSE) {
    # I prefer straight data.frames
    # but if you like tidyverse tibbles (the default with read_excel)
    # then just pass tibble = TRUE
    sheets <- readxl::excel_sheets(filename)
    x <- lapply(sheets, function(X) readxl::read_excel(filename, sheet = X))
    if(!tibble) x <- lapply(x, as.data.frame)
    names(x) <- sheets
    x
}
```


```{r}
txdf = ensembldb::transcripts(EnsDb.Hsapiens.v86, return.type="DataFrame")
genedf = ensembldb::genes(EnsDb.Hsapiens.v86, return.type="DataFrame") %>% 
  as.data.frame() %>% 
  dplyr::select(gene_name, gene_id)
tx2gene = as.data.frame(txdf[,c("tx_id","gene_id", "tx_biotype")])

 

tx2gene %<>% 
  group_by(gene_id) %>%
  dplyr::count(gene_id) %>% 
  as.data.frame() %>% 
  set_colnames(c("gene_id", "tx_n")) %>%
   left_join(genedf, by = "gene_id")
  
  
```

```{r}
PALETTE_TTYPE = c(
    "Lung"= "#007fff",
    "Prostate"= "#665d1e",
    "Stomach"= "#ffbf00",
    "Central Nervous System"= "#fbceb1",
    "Skin"= "#ff033e",
    "Bladder"= "#ab274f",
    "Haematopoietic and Lymphoid"= "#d5e6f7",
    "Kidney"= "#7cb9e8",
    "Thyroid"= "#efdecd",
    "Soft Tissue"= "#8db600",
    "Head and Neck"= "#e9d66b",
    "Ovary"= "#b284be",
    "Bone"= "#b2beb5",
    "Endometrium"= "#10b36f",
    "Breast"= "#6e7f80",
    "Pancreas"= "#ff7e00",
    "Peripheral Nervous System"= "#87a96b",
    "Cervix"= "#c9ffe5",
    "Large Intestine"= "#9f2b68",
    "Liver"= "#00ffff",
    "Vulva"= "#008000",
    "Esophagus"= "#cd9575",
    "Biliary Tract"= "#72a0c1",
    "Other tissue"= "#a32638",
    "Small Intestine"= "#9966cc",
    "Placenta"= "#f19cbb",
    "Testis"= "#e32636",
    "Adrenal Gland"= "#3b7a57",
    "Other"= "whitesmoke")
```



# Figure 1 
mRNA-protein correlations across tissues and cancer related pathways. 

## Figure 1A

```{r}
correlation_data_tissue = read.csv("output/correlations-datasets/correlation-tissue-nontransformed.csv", row.names = 1) 
```


```{r}
sample_md = read.delim("~/Documents/2023/RNA-protein-correlationx/Data/e0022_diann_051021_sample_mapping_averaged.txt")
```

```{r fig.height=10, fig.width=20, fig.cap="Distribution of mRNA-protein correlations across tissues in the ProCan-DepMapSanger dataset. Only tissues with at least 10 cell lines are shown."}
matrix_tissue = pivot_wider(correlation_data_tissue, names_from  = Tissue, values_from = Correlation)  %>% 
  as.data.frame() %>% 
  column_to_rownames("Gene") 

tissue_count =  sample_md %>% group_by(Tissue_type) %>% summarise(Count = n()) %>% as.data.frame() %>% 
  dplyr::rename(Tissue = Tissue_type)

medians_tissue = aggregate(Correlation ~ Tissue, data = correlation_data_tissue, FUN = median) %>% 
  as.data.frame() %>%
   left_join(tissue_count, by = "Tissue") %>%
    mutate(Tissue = recode(Tissue, 
                           "Breast" = "Breast",
                           "Central Nervous System" = "Central Nervous\nSystem",
                           "Head and Neck" = "Head and\nNeck",
                           "Esophagus" = "Esophagus", 
                         "Bladder" = "Bladder", 
                         "Skin" = "Skin", 
                         "Large Intestine" = "Large Intestine\n", 
                         "Kidney" = "Kidney", 
                         "Pancreas" = "Pancreas", 
                         "Soft Tissue" = "Soft Tissue", 
                         "Thyroid" = "Thyroid", 
                         "Peripheral Nervous System" = "Peripheral Nervous\nSystem", 
                         "Bone" = "Bone", 
                         "Ovary" = "Ovary", 
                         "Haematopoietic and Lymphoid" = "Haematopoietic and\nLymphoid", 
                         "Stomach" = "Stomach", 
                         "Lung" = "Lung", 
                         "Liver" = "Liver", 
                         "Cervix" = "Cervix")) %>%
     mutate(Label = paste0(Tissue, " (n=", Count, ")")) %>%
  arrange(Correlation) 



tissue_dists= correlation_data_tissue %>% 
  left_join(tissue_count, by = "Tissue") %>%
    group_by(Tissue) %>% 
  mutate(Label = recode(Tissue, 
                           "Breast" = "Breast",
                           "Central Nervous System" = "Central Nervous\nSystem",
                           "Head and Neck" = "Head and\nNeck",
                           "Esophagus" = "Esophagus", 
                         "Bladder" = "Bladder", 
                         "Skin" = "Skin", 
                         "Large Intestine" = "Large Intestine\n", 
                         "Kidney" = "Kidney", 
                         "Pancreas" = "Pancreas", 
                         "Soft Tissue" = "Soft Tissue", 
                         "Thyroid" = "Thyroid", 
                         "Peripheral Nervous System" = "Peripheral Nervous\nSystem", 
                         "Bone" = "Bone", 
                         "Ovary" = "Ovary", 
                         "Haematopoietic and Lymphoid" = "Haematopoietic and\nLymphoid", 
                         "Stomach" = "Stomach", 
                         "Lung" = "Lung", 
                         "Liver" = "Liver", 
                         "Cervix" = "Cervix")) %>%
   mutate(Label = paste0(Label, " (n=", Count, ")")) %>%
  mutate(Label = factor(Label, levels = medians_tissue$Label)) %>%
  ggplot(aes(x = fct_rev(Label), y = Correlation)) + 
  ggdist::stat_halfeye(
    aes(color = Tissue,
        fill = after_scale(lighten(color, .5))),
    adjust = .5, 
    width = .75, 
    .width = 0,
    justification = -.4, 
    point_color = NA) + theme(legend.position = "none") +
  geom_boxplot(
    aes(color = Tissue,
        color = after_scale(darken(color, .1, space = "HLS"))),
    width = .42, 
    outlier.shape = NA
  )  + 
  stat_summary(
    geom = "text",
    fun = "median",
    aes(label = round(..y.., 2),
        color = Tissue,
        color = after_scale(darken(color, .1, space = "HLS"))),
    fontface = "bold",
    size = 8,
    vjust = 10.5,
    na.rm = TRUE
  ) +
  scale_color_manual(values = PALETTE_TTYPE, guide = "none") +
  scale_fill_manual(values = PALETTE_TTYPE, guide = "none") +
  theme_bw()  + 
  labs(y="mRNA-protein correlation",
       x= "") +
  theme(axis.text.x = element_text(angle=70,  hjust = 1, size =30),
        axis.text.y = element_text(size =30),
    axis.text = element_text(color = "black", 
                                 family="serif", 
                                 size=32), 
    axis.title = element_text(color = "black", 
                                 family="serif", 
                                 size=32)) 

tissue_dists

 ggsave(filename = "output/Publication/Figures/Fig_1A.png", height =10, width =25, plot = tissue_dists, dpi = 1200)

```



## Figure 1B and Figure 1C

```{r}
cptac = read.csv("output/correlations-datasets/correlation-overall-cptac-linkedomics.csv", header=TRUE)
ccle = read.csv("output/correlations-datasets/CCLE-correlations-overall.csv", header=TRUE)
correlation_overall= read.csv("output/correlations-datasets/correlation-all-nontransformed.csv",
                            header=TRUE, row.names = 1)
```

```{r}
cptac %>% 
  mutate(Protein =  gsub("\\.\\d+$", "", Protein)) %>% 
   mutate("Gene" = mapIds(org.Hs.eg.db, keys = Protein,
                        keytype = "ENSEMBL", column = "SYMBOL", multiVals="first")) %>% 
  dplyr::rename("CPTAC"= "Correlation") %>%
  inner_join(correlation_overall %>% 
               dplyr::rename("Gene" = "Protein"), by = "Gene") %>%
  with(cor(.$Correlation, .$CPTAC, method = "spearman"))
```

```{r}
cptac %>% 
  mutate(Protein =  gsub("\\.\\d+$", "", Protein)) %>% 
   mutate("Gene" = mapIds(org.Hs.eg.db, keys = Protein,
                        keytype = "ENSEMBL", column = "SYMBOL", multiVals="first")) %>% 
  dplyr::rename("CPTAC"= "Correlation") %>%
  inner_join(ccle %>% 
               dplyr::rename("Gene" = "Protein"), by = "Gene") %>%
  with(cor(.$Correlation, .$CPTAC, method = "spearman"))
```

```{r}
ccle %>%
  dplyr::rename("CCLE"= "Correlation") %>%
  inner_join(correlation_overall,
               by = "Protein") %>%
  with(cor(.$Correlation, .$CCLE, method = "spearman"))
```

```{r}
overall_correlations_cell_lines = correlation_overall  %>%
  ggplot(aes(x = Correlation)) + 
  geom_histogram(color="white", alpha=1, fill = "#ccd5ae") + theme_bw() +
  labs(x="mRNA-protein correlation", 
       y = "Number of associations") +  annotate("text", x = -0.1, y = 600,
  label = "paste(italic(median), \" = 0.42\")", parse = TRUE, color = "#9a031e") + 
  geom_vline(xintercept = 0.42, linetype = "dashed", color = "#9a031e") +
  ggtitle("ProCan-DepMapSanger") +
     theme(legend.position = "none",
      axis.text=element_text(size=15,  family="serif", color = "black"),
      axis.title = element_text(size=16, family = "serif", color = "black")) 

```

```{r fig.height=4,fig.width=10, fig.cap="Histogram of mRNA-protein correlations across each cohort for ProCan-DepMapSanger, CPTAC and CCLE with medians indicated"}
overall_correlations_ccle = ccle   %>%
  ggplot(aes(x = Correlation)) + 
  geom_histogram(color="white", alpha=1, fill = "#ccd5ae") + theme_bw() +
  labs(x="mRNA-protein correlation", 
       y = "Number of associations") +  annotate("text", x = -0.1, y = 700,
  label = "paste(italic(median), \" = 0.46\")", parse = TRUE, color = "#9a031e") + 
  geom_vline(xintercept = 0.4555411, linetype = "dashed", color = "#9a031e") +
  ggtitle("CCLE") +
     theme(legend.position = "none",
      axis.text=element_text(size=15,  family="serif", color = "black"),
      axis.title = element_text(size=16, family = "serif", color = "black")) 


#median(cptac$Correlation,na.rm = TRUE)
overall_correlations_cptac = cptac  %>%
  ggplot(aes(x = Correlation)) + 
  geom_histogram(color="white", alpha=1, fill = "#ccd5ae") + theme_bw() +
  labs(x="mRNA-protein correlation", 
       y = "Number of associations") +  annotate("text", x = -0.2, y = 1250,
  label = "paste(italic(median), \" = 0.38\")", parse = TRUE, color = "#9a031e") + 
  geom_vline(xintercept =0.3816182, linetype = "dashed", color = "#9a031e") +
  ggtitle("CPTAC") +
     theme(legend.position = "none",
      axis.text=element_text(size=15,  family="serif", color = "black"),
      axis.title = element_text(size=16, family = "serif", color = "black")) 


ggarrange(overall_correlations_cell_lines,overall_correlations_cptac,overall_correlations_ccle, ncol= 3)




```

```{r}
 ggsave(filename = "output/Publication/Figures/Fig_1B.png", height =4, width =4, plot = overall_correlations_cell_lines, dpi = 1200)
 
 
ggsave(filename = "output/Publication/Figures/Fig_1C_left.png", height =4, width =4, plot = overall_correlations_cptac, dpi = 1200)
 
ggsave(filename = "output/Publication/Figures/Fig_1C_right.png", height =4, width =4, plot = overall_correlations_ccle, dpi = 1200)
```

## Figure 1D

Selected pathways were retrieved from Arad & Geiger (2023).

```{r}
kegg_pathways = msigdbr(species="human", category = "C2", subcategory = "CP:KEGG") %>%
  as.data.frame() %>%
  dplyr::filter(gs_description %in% c("Cell cycle", "Oxidative phosphorylation",
                                      "Ribosome", "Proteasome", "Spliceosome",
                                      "Focal adhesion", "Apoptosis") )

lists = split(kegg_pathways, kegg_pathways$gs_description)

kegg_genes = lapply(lists , function(x){
  x %<>% dplyr::select(Gene = gene_symbol)
})

pathway_genes = read_excel_allsheets("data/Pathways/Pathways-member-genes.xlsx")
pathway_genes = lapply(pathway_genes[1:10] , function(x){
  x %<>% dplyr::select(Gene)
  }
                       )


tissue_matrix = correlation_data_tissue %>%
  pivot_wider(names_from  = Tissue, values_from = Correlation)  %>%
  as.data.frame() %>%
  dplyr::select(-Protein)



correlation_per_kegg = lapply(names(kegg_genes), function(x) {
  pathways = kegg_genes[[x]]$Gene
  pathway_medians = tissue_matrix %>%
  dplyr::filter(Gene %in% pathways) %>%
  dplyr::select(-c(Gene)) %>%
  summarise_all(.funs = ~median(., na.rm = TRUE)) %>%
  mutate(Pathway = x)

  return(pathway_medians)


})

correlation_per_kegg %<>% do.call(rbind,.)




correlation_per_pathway = lapply(names(pathway_genes), function(x) {
  pathways = pathway_genes[[x]]$Gene
  pathway_medians = tissue_matrix %>%
  dplyr::filter(Gene %in% pathways) %>%
  dplyr::select(-c(Gene)) %>%
  summarise_all(.funs = ~median(., na.rm = TRUE)) %>%
  mutate(Pathway = x)

  return(pathway_medians)


})
```

```{r fig.height=11, fig.width=11, fig.cap="Boxplot showing median mRNA-protein correlations by tissue in the ProCan-DepMapSanger dataset for key biological and selected cancer-related pathways"}
plot = do.call(rbind, correlation_per_pathway) %>%
  dplyr::filter(Pathway %in% c("PI3K", "NOTCH", "WNT", "RTK RAS", "TGF-Beta",
                               "HIPPO")) %>%
  rbind(correlation_per_kegg) %>%
  mutate(Pathway = ifelse(Pathway == "Oxidative phosphorylation", "OXPHOS", Pathway)) %>%
  melt() %>%
  mutate(Pathway = factor(Pathway, levels =
                            c("Ribosome", "OXPHOS",
                              "Proteasome", "Spliceosome", "PI3K",
                              "NOTCH", "Cell cycle", "Apoptosis",
                              "WNT", "RTK RAS", "TGF-Beta", "HIPPO",
                              "Focal adhesion"))) %>%
  ggplot(aes(x=Pathway, y=value)) + geom_boxplot() +
  theme_bw() +   theme_bw() +  #geom_jitter(aes(color = variable), size=4, shape = 1, stroke = 1.5) +
  geom_jitter(
    aes(color = variable,
        #color = after_scale(darken(color, .1, space = "HLS")),
        color = after_scale(desaturate(lighten(color, .25), .4))),
    size=3, stroke = 1
  ) +
 scale_color_manual(
  values = PALETTE_TTYPE, 
  labels = c("Breast" = "Breast",
                           "Central Nervous System" = "Central Nervous\nSystem",
                           "Head and Neck" = "Head and\nNeck",
                           "Esophagus" = "Esophagus", 
                         "Bladder" = "Bladder", 
                         "Skin" = "Skin", 
                         "Large Intestine" = "Large Intestine", 
                         "Kidney" = "Kidney", 
                         "Pancreas" = "Pancreas", 
                         "Soft Tissue" = "Soft Tissue", 
                         "Thyroid" = "Thyroid", 
                         "Peripheral Nervous System" = "Peripheral Nervous\nSystem", 
                         "Bone" = "Bone", 
                         "Ovary" = "Ovary", 
                         "Haematopoietic and Lymphoid" = "Haematopoietic\nand Lymphoid", 
                         "Stomach" = "Stomach", 
                         "Lung" = "Lung", 
                         "Liver" = "Liver", 
                         "Cervix" = "Cervix"))  + # Split into 6 rows
  labs(y= "Median mRNA-protein correlation",
       color = "Tissue") + theme(legend.position = "top",
        legend.spacing.y = unit(0, "mm"),
        panel.border = element_rect(colour = "black", fill=NA),
        legend.background = element_blank(),
        legend.box.background = element_rect(colour = "black"),
    legend.text = element_text(family = "serif", size=18),
    legend.title = element_text(family = "serif", size=18)) +
  theme(axis.text.x = element_text( color = "black", size =25, family = "serif", hjust = 1),
        axis.text.y = element_text(color = "black", size =25, family = "serif"),
        axis.title = element_text(color = "black", size =25, family = "serif"),
        legend.position = "right", legend.spacing.x = unit(0.8, 'cm')) + xlab("") + coord_flip()
plot

ggsave(filename = "output/Publication/Figures/Fig_1D.png", height =11, width =11, plot = plot, dpi = 1200)


```

## Figure 1E 


```{r fig.height=8, fig.width=5, fig.cap="Correlation of median mRNA-protein correlations per tissue for CPTAC with ProCan-DepMapSanger (top), and with CCLE (bottom), for pathways in Figure 1d"}
pathways_across_datasets = read.table("output/Other/Pathways_median_all_datasets.tsv", sep = "\t", header =TRUE)



median_pathway <- pathways_across_datasets %>%
  group_by(Pathway, Dataset) %>%  # Group by Pathway and Dataset
  summarise(Median_Correlation = median(`Median.Correlation`, na.rm = TRUE), .groups = "drop") %>% 
  as.data.frame() %>% 
  pivot_wider(names_from = Dataset, values_from = Median_Correlation) %>% as.data.frame() %>% 
  dplyr::rename(ProCan = `ProCan-DepMapSanger`)

library(ggpubr)



pathways_cor_1 = median_pathway %>%

 ggplot(aes(x = `CPTAC`, y = ProCan)) +
  geom_point(size =3) + geom_smooth(method = "lm", color = "blue", fill = "lightgray", se = TRUE) +
   theme_bw() + xlab("CPTAC") +
  ylab("ProCan-DepMapSanger") +
  annotate("text", x = 0.22, y = 0.5, size = 5,
  label = "paste(italic(r), \" = 0.85, \", italic(p), \" = 0.00034\")", 
  parse = TRUE, color = "black") +
  theme(axis.text.x = element_text(size = 25, family = "serif", color = "black"),
            axis.text.y = element_text(size = 25, family = "serif", color = "black"),
        axis.title = element_text(size=25, family = "serif"),
         legend.box.background = element_rect(color = "black"),
          legend.text = element_text(family = "serif"),
          legend.title = element_text( family = "serif"),
           plot.title = element_text(family = "serif", size =20),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line( size=.1 ),
              legend.position = "top",
        strip.text.y = element_text(
        size = 16, face = "bold.italic", family = "serif"
        ),
         strip.text.x = element_text(
        size = 16, family = "serif"
        )) 
 

 
ggsave(filename = "output/Publication/Figures/Fig_1E_upper.png",height = 5, width = 8, plot =pathways_cor_1, dpi = 1200 )


pathways_cor_2 = median_pathway %>% 
 ggplot(aes(x = `CPTAC`, y = CCLE)) +
  geom_point(size =3) + geom_smooth(method = "lm", color = "blue", fill = "lightgray", se = TRUE) +
   theme_bw() + xlab("CPTAC") +
  ylab("CCLE") +
  annotate("text", x = 0.22, y = 0.57, size = 5,
  label = "paste(italic(r), \" = 0.74, \", italic(p), \" = 0.0053\")", 
  parse = TRUE, color = "black") +
  theme(axis.text.x = element_text(size = 25, family = "serif", color = "black"),
            axis.text.y = element_text(size = 25, family = "serif", color = "black"),
        axis.title = element_text(size=25, family = "serif"),
         legend.box.background = element_rect(color = "black"),
          legend.text = element_text(family = "serif"),
          legend.title = element_text( family = "serif"),
           plot.title = element_text(family = "serif", size =20),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line( size=.1 ),
              legend.position = "top",
        strip.text.y = element_text(
        size = 16, face = "bold.italic", family = "serif"
        ),
         strip.text.x = element_text(
        size = 16, family = "serif"
        )) 
 

 ggsave(filename = "output/Publication/Figures/Fig_1E_lower.png",height = 5, width = 8, plot =pathways_cor_2, dpi = 1200 )

 
 ggarrange(pathways_cor_1, pathways_cor_2, nrow=2)


```

# Figure 2 

mRNA-protein correlations of cancer driver genes from OncoKB.

## Figure 2A


```{r}
drivers = read.delim(here::here("data/OncoKb/cancerGeneList.tsv"), fill=TRUE, check.names = FALSE)
```

```{r}
correlation_data_tissue = read.csv("output/correlations-datasets/correlation-tissue-nontransformed.csv", row.names = 1) 
matrix_tissue = pivot_wider(correlation_data_tissue, names_from  = Tissue, values_from = Correlation)  %>% 
  as.data.frame() %>% 
  column_to_rownames("Gene") 


driver_matrix = matrix_tissue %>%
  rownames_to_column("Gene") %>%
  dplyr::filter(Gene %in% drivers$`Hugo Symbol`) %>%
  dplyr::select(-Protein) %>%
  column_to_rownames("Gene")


tmp=rowSums(is.na(driver_matrix) )

tmp=tmp[tmp<5]


matrix_drivers=driver_matrix[names(tmp),]

matrix_drivers %<>% as.matrix()


#km = kmeans(na.omit(matrix_drivers), 10)
```



```{r fig.height=6, fig.width=6, fig.cap=" Heatmap of mRNA-protein correlations across tissue types for cancer driver genes from OncoKB. Clusters identidied by hierachical clustering using Euclidean distance."}
heatmap_viz= matrix_drivers %>%
  #  drop_na() %>%

pheatmap::pheatmap( border_color = "white",
         breaks = seq(-1,1, length.out=100),
         color = colorRampPalette(c("#2F124b", "#6e74b4", "#f9f9f9",
                                    "#fd9675", "#Ea2a5f"))(100),
         na_col = "darkgrey", cutree_rows    = 5, show_rownames = FALSE, fontsize_col=12)


ggsave(plot = heatmap_viz, filename = "output/Publication/Figures/Fig_2A.png",
      height = 8, width = 8, dpi = 1200)


```


## Figure 2B 


```{r}
matrix_tissue = pivot_wider(correlation_data_tissue, names_from  = Tissue, values_from = Correlation)  %>%
  as.data.frame() %>%
  column_to_rownames("Gene") %>%
  dplyr::select(-Protein)
```

```{r fig.height= 5, fig.width=5, fig.cap="Violin plot showing mRNA-protein correlations of cancer driver and non-driver genes with median and standard deviation indicated. **** indicates P < 0.0001 by unpaired Student’s t-test."}

drivers = read.delim(here::here("data/OncoKb/cancerGeneList.tsv"), fill=TRUE, check.names = FALSE)
tmp=rowSums(is.na(matrix_tissue ) )

tmp=tmp[tmp<5]

matrix_filtered=matrix_tissue[names(tmp),]

n_counts <- correlation_overall %>%
  dplyr::filter(Protein %in% rownames(matrix_filtered)) %>%
  mutate(Gene = ifelse(Protein %in% drivers$`Hugo Symbol`, "Cancer Driver", "Non-Driver")) %>%
  ungroup() %>%  # Remove any grouping
  mutate(Gene = as.character(Gene)) %>%  # Convert to character
  group_by(Gene) %>%  # Group by Gene
  summarise(n = n(), .groups = "drop") %>%  # Count occurrences
  mutate(Label = paste0(Gene, "\n(n = ", n, ")"))  # Create formatted labels


correlation_overall %>%
  dplyr::filter(Protein %in% rownames(matrix_filtered)) %>%
  mutate(Gene = ifelse(Protein %in% drivers$`Hugo Symbol`,
                       "Cancer Driver", "Non-Driver")) %>%
  mutate(Gene = factor(Gene, levels = c("Non-Driver", "Cancer Driver"))) %>%
  dplyr::group_by(Gene) %>%
  dplyr::summarise(
    median_correlation = median(Correlation, na.rm = TRUE),
    n = dplyr::n()
  ) 


df <- correlation_overall %>%
  dplyr::filter(Protein %in% rownames(matrix_filtered)) %>%
  mutate(Gene = ifelse(Protein %in% drivers$`Hugo Symbol`,
                       "Cancer Driver", "Non-Driver")) %>%
  mutate(Gene = factor(Gene, levels = c("Non-Driver", "Cancer Driver")))

t.test(Correlation ~ Gene,
       data = df,
#       paired = FALSE,
       var.equal = TRUE)


drivers_v_nondrivers <- correlation_overall %>%
  dplyr::filter(Protein %in% rownames(matrix_filtered)) %>%
  mutate(Gene = ifelse(Protein %in% drivers$`Hugo Symbol`, "Cancer Driver", "Non-Driver")) %>%
  mutate(Gene = factor(Gene, levels = c("Non-Driver", "Cancer Driver"))) %>%
  ggplot(aes(x = Gene, y = Correlation, fill = Gene)) +

  # Violin plot
  geom_violin(alpha = 0.6) +

  # Boxplot
  geom_boxplot(width = 0.2, fill = "white", outlier.shape = NA) +

  # Add significance test
  geom_signif(comparisons = list(c("Non-Driver", "Cancer Driver")),
              map_signif_level = TRUE,
               annotations = "**") +



  # Custom colors
  scale_fill_manual(values = c("Cancer Driver" = "#f08080", "Non-Driver" = "#adb5bd")) +
    scale_x_discrete(labels = setNames(n_counts$Label, n_counts$Gene)) +

  # Labels and theme
  labs(y = "mRNA-protein correlation", x = "") +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 20, family = "serif", color = "black"),
    axis.text.y = element_text(size = 20, family = "serif", color = "black"),
    axis.title = element_text(size = 20, family = "serif"),
    legend.box.background = element_rect(color = "black"),
    legend.text = element_text(family = "serif", size=15),
    legend.title = element_text(family = "serif", size=15),
    plot.title = element_text(family = "serif", size = 20),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(size = 0.1),
    legend.position = "top",
    strip.text.y = element_text(size = 10, face = "bold.italic", family = "serif"),
    strip.text.x = element_text(size = 10, family = "serif")
  )

drivers_v_nondrivers
ggsave(filename = "output/Publication/Figures/Fig_2B.png",height = 6.5, width = 6, plot =drivers_v_nondrivers, dpi = 1200 )
```


## Figure 2C


```{r fig.height=10, fig.width=7, fig.cap="Distribution of mRNA-protein correlations for genes in each cluster identified in heatmap in Figure 2A"}
matrix <- matrix_drivers[c(heatmap_viz$tree_row[["order"]]),heatmap_viz$tree_col[["order"]]]
clusters= cutree(heatmap_viz$tree_row, k=5)[heatmap_viz$tree_row[["order"]]] %>%
  as.data.frame() %>%
  set_colnames("Cluster")


clusters$clusters <- NA

clusters$clusters[clusters$Cluster == 5] <- 1
clusters$clusters[clusters$Cluster == 1] <- 2
clusters$clusters[clusters$Cluster == 4] <- 3
clusters$clusters[clusters$Cluster == 3] <- 4
clusters$clusters[clusters$Cluster == 2] <- 5

# 
# clusters %>%
#    as.data.frame() %>%
#    rownames_to_column("Gene") %>%
#    inner_join(matrix_drivers %>% melt() %>%
#                 dplyr::rename("Gene" = "Var1",  "Correlation" = "value" ), by = "Gene") %>%
#    group_by(Cluster) %>%
#    mutate(median= median(Correlation, na.rm =TRUE)) %>%
#    distinct("median", .keep_all = T)


clusters_drivers = clusters %>%
  as.data.frame() %>%
  rownames_to_column("Gene") %>%
  inner_join(matrix_drivers %>% melt() %>%
               dplyr::rename("Gene" = "Var1",  "Correlation" = "value" ), by = "Gene") %>%
  group_by(clusters) %>%
  mutate(median= median(Correlation, na.rm =TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x=Correlation)) +
   geom_histogram(color="black", alpha=0.4, fill = "#6e74b4") + theme_bw() +
  facet_grid(rows = vars(clusters), scales = "free_y") + ylab("Number of Proteins") +
  geom_vline(aes(xintercept = median), linetype = "dashed", color = "#9a031e", size =1) +
    theme(legend.position = "top",
                          legend.title = element_blank(),
        legend.spacing.y = unit(0, "mm"),
        panel.border = element_rect(colour = "black", fill=NA),
        legend.background = element_blank(),
        legend.box.background = element_rect(colour = "black")) +
  theme(axis.text.x = element_text( color = "black", size =16, family = "serif", hjust = 1),
        axis.text.y = element_text(color = "black", size =16, family = "serif"),
        axis.title = element_text(color = "black", size =16, family = "serif"),
        legend.position = "top") + xlab("mRNA-protein correlation")

clusters_drivers

clusters %>%
    as.data.frame() %>%
    rownames_to_column("Gene") %>%
    inner_join(matrix_drivers %>% reshape2::melt() %>%
                 dplyr::rename("Gene" = "Var1",  "Correlation" = "value" ), by = "Gene") %>%
    group_by(clusters) %>%
    mutate(median= median(Correlation, na.rm =TRUE)) %>%
    distinct("median", .keep_all = T)


 # ggsave(plot = clusters_drivers, filename = "output/Publication/Figures/Figure_2b_update.png",
#        height = 8, width = 7, dpi = 1200)

# 
 cluster_full = correlation_data_tissue %>%
   left_join(clusters %>%
               rownames_to_column("Gene"), by = "Gene")
write.csv(cluster_full, "output/Publication/Data/clusters.csv")
```

### Gene ontologies

GO terms for each cluster were identified using the clusterProfiler and saved in a CSV. The top terms were combined with the distribution plot in Adobe Illustrator.


```{r}
universe = correlation_data_tissue$Protein %>%
  as.data.frame() %>%
  set_colnames("Proteins") %>%
  mutate("Protein_name" = gsub(";.*", "", Proteins)) %>%
  mutate("Gene" = mapIds(org.Hs.eg.db, keys = Protein_name,
                         keytype = "UNIPROT", column = "SYMBOL", multiVals="first")) %>%
  mutate("Entrez" = mapIds(org.Hs.eg.db, keys = Protein_name,
                         keytype = "UNIPROT", column = "ENTREZID", multiVals="first"))
```

```{r}
library(clusterProfiler)
gene_ontology = list()
for (clus in unique(clusters$clusters)) {
  df =  clusters %>%
    dplyr::filter(clusters == clus) %>%
    rownames_to_column("Gene") %>%
      mutate("Entrez" = mapIds(org.Hs.eg.db, keys = Gene,
                         keytype = "SYMBOL", column = "ENTREZID", multiVals="first"))

  ego = enrichGO(gene          = df$Entrez,
                universe      = universe$Entrez,
                OrgDb         = org.Hs.eg.db,
                ont           = "ALL",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
        readable      = TRUE)

  gene_ontology[[paste0("cluster", clus)]] = head(ego, n=100) %>%
    as.data.frame() %>%
    dplyr::select( Description, "p.adjust", "qvalue", geneID, Count) %>%
    mutate(Cluster = paste("Cluster", clus))

}


gene_ontology_top = lapply(gene_ontology, function(x){
  ont= x %>%
    arrange(qvalue) %>%
    head(200)
  return(ont)
})

library(openxlsx)
write.xlsx(gene_ontology_top, file = "output/Publication/Data/Gene-ontology-Figure2b.xlsx")
```

```{r}
gene_ontology_top$cluster1 %>% 
  DT::datatable(caption = "Gene ontology for Cluster 1")
```

```{r}
gene_ontology_top$cluster2 %>% 
  DT::datatable(caption = "Gene ontology for Cluster 2")
```

```{r}
gene_ontology_top$cluster3 %>% 
  DT::datatable(caption = "Gene ontology for Cluster 3")
```

```{r}
gene_ontology_top$cluster4 %>% 
  DT::datatable(caption = "Gene ontology for Cluster 4")
```

```{r}
gene_ontology_top$cluster5 %>% 
  DT::datatable(caption = "Gene ontology for Cluster 5")
```


```{r}

cluster_full = clusters %>%
  as.data.frame() %>%
  rownames_to_column("Gene") %>%
  inner_join(matrix_drivers %>% melt() %>%
               dplyr::rename("Gene" = "Var1",  "Correlation" = "value" ), by = "Gene") %>%
  dplyr::select(-Cluster)

write.csv(cluster_full, "output/Publication/Data/clusters.csv")
```


# Figure 3 

Integration of protein reproducibility ranks and gene-specific mRNA-protein correlations. Protein reproducibility ranks were retrieved from [Upadhya et al (2022)](https://www.cell.com/cell-reports-methods/fulltext/S2667-2375(22)00170-9?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS2667237522001709%3Fshowall%3Dtrue). 
## Figure 3A


```{r}
correlation_overall= read.csv("output/correlations-datasets/correlation-all-nontransformed.csv",
                            header=TRUE, row.names = 1) %>%
  drop_na()

ranks = read_excel("data/Ranks/1-s2.0-S2667237522001709-mmc3.xlsx", sheet=3) %>%
  as.data.frame() %>%
  dplyr::rename("Gene" = "...1") %>%
  mutate(Bin = case_when(
    between(`Aggregated Reproducibility Rank`, 0, 0.2) ~ "0 - 0.2",
    between(`Aggregated Reproducibility Rank`, 0.2, 0.4) ~ "0.2 - 0.4",
    between(`Aggregated Reproducibility Rank`, 0.4, 0.6) ~ "0.4 - 0.6",
    between(`Aggregated Reproducibility Rank`, 0.6, 0.8) ~ "0.6 - 0.8",
    between(`Aggregated Reproducibility Rank`, 0.8, 1) ~ "0.8 - 1"
  )) %>%
  mutate(Bin = factor(Bin, levels = c("-1 - -0.8","-0.8 - -0.6",
                                  "-0.6 - -0.4", "-0.4 - -0.2",
                                  "-0.2 - 0","0 - 0.2",
                                  "0.2 - 0.4", "0.4 - 0.6",
                                  "0.6 - 0.8",
                                  "0.8 - 1")))


ranks_plot = correlation_overall %>%
  dplyr::rename("Gene" = "Protein") %>%
  inner_join(ranks, by = "Gene") %>%
  drop_na("Aggregated Reproducibility Rank") %>%
    ggplot(aes(x=Bin, y = `Correlation`)) + geom_boxplot(fill="#be95c4") + theme_bw() +
  labs(x="Aggregated protein reproducibility rank\n(binned)",
       y="mRNA-protein correlation") +
  #annotate("text", x = 1, y = 0.9, size=6,
  #label = "paste(italic(r), \" = 0.48\")", parse = TRUE, color = "black") +
  theme(axis.text.x = element_text( color = "black", size =12),
        axis.text.y = element_text(color = "black", size =12),
        axis.title = element_text(color = "black", size =15)) +
    theme(
        axis.text.y = element_text(size = 20, family = "serif", color = "black"),
         axis.text.x = element_text(size = 20, family = "serif", color = "black"),
        axis.title = element_text(size=20, family = "serif"),
         legend.box.background = element_rect(color = "black"),
          legend.text = element_text(family = "serif"),
          legend.title = element_text( family = "serif"),
           plot.title = element_text(family = "serif", size =20),
        panel.grid.major.x = element_blank(),
              legend.position = "top") +  ggtitle("ProCan-DepMapSanger")


correlation_overall %>%
  dplyr::rename("Gene" = "Protein") %>%
  inner_join(ranks, by = "Gene") %>%
  drop_na("Aggregated Reproducibility Rank") %>%
  with(cor.test(.$Correlation, .$`Aggregated Reproducibility Rank`, method = "spearman"))

ggsave(filename = "output/Publication/Figures/Fig_3A_upper.png", height =3.5, width =7, plot = ranks_plot, dpi = 1200)
```


```{r}
library(ggpubr)

correlation_ranks = correlation_overall %>%
  dplyr::rename("Gene" = "Protein") %>%
  inner_join(ranks, by = "Gene") %>%
  drop_na("Aggregated Reproducibility Rank") %>%
 ggplot(aes(x = `Aggregated Reproducibility Rank`, y = Correlation)) +
  geom_point(size =3) + geom_smooth(method = "lm", color = "blue", fill = "lightgray", se = TRUE) +
    theme_bw() + ylab("mRNA-protein correlation") +
  xlab("Aggregated protein reproducibility rank") +
  annotate("text", x = 0.2, y = 0.85, size = 6,
  label = "paste(italic(r), \" = 0.49, \", italic(p), \" < 2.2e-16\")",
  parse = TRUE, color = "blue") +
  theme(axis.text.x = element_text(size = 20, family = "serif", color = "black"),
            axis.text.y = element_text(size = 20, family = "serif", color = "black"),
        axis.title = element_text(size=20, family = "serif"),
         legend.box.background = element_rect(color = "black"),
          legend.text = element_text(family = "serif"),
          legend.title = element_text( family = "serif"),
           plot.title = element_text(family = "serif", size =20),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line( size=.1 ),
              legend.position = "top",
        strip.text.y = element_text(
        size = 16, face = "bold.italic", family = "serif"
        ),
         strip.text.x = element_text(
        size = 16, family = "serif"
        ))




ggsave(filename = "output/Publication/Figures/Fig_3A_lower.png", height =3.5, width =7, plot = correlation_ranks, dpi = 1200)
```


```{r fig.height=4, fig.width=4, fig.cap="Association between aggregated protein reproducibility ranks from Upadhya et al 22 and mRNA-protein correlations in the ProCan-DepMapSanger dataset."}
ggarrange(ranks_plot, correlation_ranks, nrow = 2)
```

## Figure 3B

```{r}
drivers = read.delim(here::here("data/OncoKb/cancerGeneList.tsv"), fill=TRUE, check.names = FALSE)


ranks_all_1 = correlation_overall %>%
dplyr::rename("Gene"= "Protein") %>%
  inner_join(ranks, by = "Gene")
```

```{r fig.height=6, fig.width=6, fig.cap=" Scatter plot of mRNA-protein correlation against aggregated protein reproducibility rank, with cancer drivers indicated by triangle data points."}
library(ggpubr)
drivers_of_interest_1 = c("CTNNB1", "EGFR","IDH2","RAC2","LCP1","ALDH2", "CPS1", "KMT2D", "DPYD")
drivers_of_interest = c("CTNNB1", "EGFR","IDH2","ALDH2","KMT2D")

ranks_all_1 %<>%
  mutate(Drivers = ifelse(Gene %in% drivers$`Hugo Symbol`, "Cancer Driver", "Non-Driver")) %>%
  dplyr::select(-c(`Ovarian Reproducibility Rank`, `Colon Reproducibility Rank`, `CCLE Reproducibility Rank`)) %>%
  dplyr::select(Gene, everything())


ranks_v_correlation = ranks_all_1 %>%
  mutate(Drivers = ifelse(Gene %in% drivers$`Hugo Symbol`, "Cancer Driver", "Non-Driver"),
         Gene_label = ifelse(Gene %in% drivers_of_interest, Gene, NA)) %>%


    ggscatter(.,x="Correlation", y ="Aggregated Reproducibility Rank", add = "reg.line",
               color = "Drivers",size = "Drivers", shape = "Drivers", add.params = list(color = "blue",
                                                 fill = "lightgray"), alpha = "Drivers",  cor.method = "spearman", cor.coef.size = 5.5) +
    theme_bw() + xlab("mRNA-protein correlation") +
  ylab("Aggregated protein reproducibility rank") +
  annotate("text", x = -0.2, y = 0.95, size = 6,
  label = "paste(italic(r), \" = 0.49, \", italic(p), \" < 2.2e-16\")",
  parse = TRUE, color = "blue")+


  geom_vline(xintercept = 0.5, size = 0.7, lty = "dashed", color = "grey60") + theme(legend.position = "none") +
  geom_hline(yintercept = 0.4 , size = 0.7, lty = "dashed", color = "grey60") +
  theme(axis.text.x = element_text(size = 26, family = "serif", color = "black"),
            axis.text.y = element_text(size = 26, family = "serif", color = "black"),
        axis.title = element_text(size=26, family = "serif"),
         legend.box.background = element_rect(color = "black"),
          legend.text = element_text(family = "serif", size =22),
          legend.title = element_text( family = "serif", size =22),
           plot.title = element_text(family = "serif", size =23),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line( size=.1 ),
              legend.position = "top",
        strip.text.y = element_text(
        size = 16, face = "bold.italic", family = "serif"
        ),
         strip.text.x = element_text(
        size = 16, family = "serif"
        )) + scale_shape_manual(values = c("Cancer Driver" = 17,
                                           "CTNNB1" = 17,
                                           "CPS1" = 17,
                                           "KMT2D" = 17,
                                           "MYH11" = 17,
                                            "DPYD" = 17,
                                           "Non-Driver" = 19)) +
  scale_size_manual(values = c("Cancer Driver" = 6,
                                           "CTNNB1" = 6,
                                           "MYH11" = 6,
                                           "CPS1" = 6,
                                           "KMT2D" = 6,
                                            "DPYD" = 6,
                                           "Non-Driver" = 3)) +
  scale_color_manual(values = c("Non-Driver" = "lightgrey",
                                "Cancer Driver" = "#ad2831",
                                "CTNNB1" = "#ad2831",
                                           "CPS1" = "#ad2831",
                                           "KMT2D" = "#ad2831",
                                "DPYD" = "#ad2831",
                                "MYH11" = "#ad2831")) +
  theme(legend.position = "top") +
    scale_alpha_manual(values = c("Non-Driver" = 0.45,
                                "Cancer Driver" = 1,
                                "CTNNB1" = 1,
                                           "TPR" = 1,
                                           "EGFR" = 1,
                                           "LPP" = 1,
                                           "MTOR" = 1,
                                           "IDH2" = 1,
                                "MYH11" = 1)) +
  geom_text(
    data = . %>% dplyr::filter(!is.na(Gene_label)),  # Filter data to only include rows with labels
    aes(label = Gene_label),  # Use the Gene_label column for labels
    vjust = -0.3,
    hjust = -0.001,
    color = "black",
    size = 6.5,
    family = "serif",
  ) + labs(alpha = "Gene",
           shape = "Gene",
           colour = "Gene",
           size = "Gene")

ranks_v_correlation


ggsave(filename = "output/Publication/Figures/Fig_3B.png", height =9, width = 10, plot = ranks_v_correlation, dpi = 1200)
```

## Figure 3C

```{r}
correlation_data_tissue = read.csv("output/correlations-datasets/correlation-tissue-nontransformed.csv", row.names = 1)
ranks_all = correlation_data_tissue %>%
  inner_join(ranks, by = "Gene")
```

```{r}
cr_correlations <- read.csv("output/correlations-datasets/correlation-tissue-cptac-linkedomics.csv", row.names = 1) %>%
  mutate(Protein =  gsub("\\.\\d+$", "", Protein)) %>%
   mutate("Gene" = mapIds(org.Hs.eg.db, keys = Protein,
                        keytype = "ENSEMBL", column = "SYMBOL", multiVals="first"))

ccle = read.csv("output/correlations-datasets/CCLE-correlations-tissue.csv", row.names = 1)
```



* ALDH2

```{r}
ALDH2  = ranks_all %>%
  ggplot(aes(Correlation)) +
   geom_density(data=ranks_all, color="darkgrey", size = 1) +
      geom_density(data=dplyr::filter(cr_correlations, Gene == "ALDH2"), color="#e3a3b3", size = 1.5, lty ="dashed", alpha = 0.5) +
   geom_density(data=dplyr::filter(ccle, Protein == "ALDH2"), color="#6e0b24", size = 1.5, lty ="dotted", alpha =0.5) +
  geom_density(data=dplyr::filter(ranks_all, Gene == "ALDH2"), color="#fb6f92", size = 2) +


  theme_classic() + labs(x= "mRNA-protein correlation",
                    y= "Density",
                    title = "ALDH2") +
theme(
  axis.text = element_text(color = "black", size = 25, family = "serif", face = "plain"),
  axis.title = element_text(color = "black", size = 25, family = "serif", face = "plain"),
  title = element_text(color = "black", size = 25, family = "serif", face = "bold")
)

ggsave(plot = ALDH2, filename = "output/Publication/Figures/ALDH2-dist.png", height = 5, width =7, dpi=1200)
```

* IDH2

```{r}
idh2 = ranks_all %>%
  ggplot(aes(Correlation)) +
   geom_density(data=ranks_all, color="lightgray", size = 1.5) +
        geom_density(data=dplyr::filter(ccle, Protein == "IDH2"), color="#6e0b24", size = 1.5, lty ="dotted") +
          geom_density(data=dplyr::filter(cr_correlations, Gene == "IDH2"), color="#e3a3b3", size = 1.5, lty ="dashed") +
  geom_density(data=dplyr::filter(ranks_all, Gene == "IDH2"), color="#fb6f92", size = 2) +


  theme_classic() + labs(x= "mRNA-protein correlation",
                    y= "Density",
                    title = "IDH2") +
  theme(legend.position = "top",
                          legend.title = element_blank(),
        legend.spacing.y = unit(0, "mm")) +
theme(
  axis.text = element_text(color = "black", size = 25, family = "serif", face = "plain"),
  axis.title = element_text(color = "black", size = 25, family = "serif", face = "plain"),
  title = element_text(color = "black", size = 25, family = "serif", face = "bold")
)

ggsave(filename = "output/Publication/Figures/IDH2_distribution.png", height =5, width =7, plot = idh2, dpi = 1200)
```


* EGFR

```{r}
egfr = ranks_all %>%
  ggplot(aes(Correlation)) +
   geom_density(data=ranks_all, color="lightgray", size = 1) +
        geom_density(data=dplyr::filter(cr_correlations, Gene == "EGFR"), color="#e3a3b3", size = 1.5, lty ="dashed") +
   geom_density(data=dplyr::filter(ccle, Protein == "EGFR"), color="#6e0b24", size = 1.5, lty ="dotted") +
  geom_density(data=dplyr::filter(ranks_all, Gene == "EGFR"), color="#fb6f92", size = 2) +


  theme_classic() + labs(x= "mRNA-protein correlation",
                    y= "Density",
                    title = "EGFR") +
    theme(legend.position = "top",
                          legend.title = element_blank(),
        legend.spacing.y = unit(0, "mm")) +
theme(
  axis.text = element_text(color = "black", size = 25, family = "serif", face = "plain"),
  axis.title = element_text(color = "black", size = 25, family = "serif", face = "plain"),
  title = element_text(color = "black", size = 25, family = "serif", face = "bold")
)


ggsave(filename = "output/Publication/Figures/EGFR_distribution.png", height =5, width =7, plot = egfr, dpi = 1200)

```

* CTNNB1

```{r}
ctnnb1 = ranks_all %>%
  ggplot(aes(Correlation)) +
   geom_density(data=ranks_all, color="darkgrey", size = 1) +
      geom_density(data=dplyr::filter(cr_correlations, Gene == "CTNNB1"), color="#e3a3b3", size = 1.5, lty ="dashed", alpha = 0.5) +
   geom_density(data=dplyr::filter(ccle, Protein == "CTNNB1"), color="#6e0b24", size = 1.5, lty ="dotted", alpha =0.5) +
  geom_density(data=dplyr::filter(ranks_all, Gene == "CTNNB1"), color="#fb6f92", size = 2) +


  theme_classic() + labs(x= "mRNA-protein correlation",
                    y= "Density",
                    title = "CTNNB1") +
theme(
  axis.text = element_text(color = "black", size = 25, family = "serif", face = "plain"),
  axis.title = element_text(color = "black", size = 25, family = "serif", face = "plain"),
  title = element_text(color = "black", size = 25, family = "serif", face = "bold")
)


ggsave(plot = ctnnb1, filename = "output/Publication/Figures/CTNNB1-dist.png", height = 5, width =7, dpi=1200)
```

* KMT2D

```{r}
kmt2d  = ranks_all %>%
  ggplot(aes(Correlation)) +
   geom_density(data=ranks_all, color="darkgrey", size = 1) +
      geom_density(data=dplyr::filter(cr_correlations, Gene == "KMT2D"), color="#e3a3b3", size = 1.5, lty ="dashed", alpha = 0.5) +
   geom_density(data=dplyr::filter(ccle, Protein == "KMT2D"), color="#6e0b24", size = 1.5, lty ="dotted", alpha =0.5) +
  geom_density(data=dplyr::filter(ranks_all, Gene == "KMT2D"), color="#fb6f92", size = 2) +


  theme_classic() + labs(x= "mRNA-protein correlation",
                    y= "Density",
                    title = "KMT2D") +
theme(
  axis.text = element_text(color = "black", size = 25, family = "serif", face = "plain"),
  axis.title = element_text(color = "black", size = 25, family = "serif", face = "plain"),
  title = element_text(color = "black", size = 25, family = "serif", face = "bold")
)



ggsave(plot = kmt2d, filename = "output/Publication/Figures/KMT2D-dist.png", height = 5, width =7, dpi=1200)
```





```{r fig.height=15, fig.width=10, fig.cap="For each gene, plots show the distribution of mRNA-protein correlations across tissues in the ProCan-DepMapSanger (pink solid line), Clinical Proteomic Tumor Analysis Consortium (CPTAC; light dashed line) and Cancer Cell Line Encyclopedia (CCLE; dark dashed line) datasets. Background (grey solid line) indicates the median mRNA-protein correlation from all genes in the ProCan-DepMapSanger dataset." }
ggarrange(ALDH2,idh2 ,
          egfr, ctnnb1,
          kmt2d, ncol = 2, nrow = 3)
```



## Figure 3D


```{r fig.height=10, fig.width=5, fig.cap="mRNA-protein correlation of CTNNB1 for each tissue type. "}
correlation_data_tissue = read.csv("output/correlations-datasets/correlation-tissue-nontransformed.csv", row.names = 1)


ctnn1_tissues = correlation_data_tissue %>%
  dplyr::filter(Gene == "CTNNB1") %>%
  ggplot(aes(x= reorder(Tissue, Correlation), y = Correlation)) + geom_bar(stat= "identity", fill= "red") + geom_hline(yintercept =  0.103 , size = 0.7, lty = "dashed", color = "blue") +

  coord_flip() +
  theme_bw() +    ggtitle("CTNNB1") + theme(axis.text.x = element_text(size = 20, family = "serif", color = "black"),
            axis.text.y = element_text(size = 24, family = "serif", color = "black"),
        axis.title = element_text(size=24, family = "serif"),
         legend.box.background = element_rect(color = "black"),
          legend.text = element_text(family = "serif"),
          legend.title = element_text( family = "serif"),
           plot.title = element_text(family = "serif", size =20),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line( size=.1 ),
              legend.position = "none",
        strip.text.y = element_text(
        size = 10, face = "bold.italic", family = "serif"
        ),
         strip.text.x = element_text(
        size = 10, family = "serif"
        )) + labs(x= "", y="Spearman Correlation") +
  annotate("text", x = 4, y = 0.16,
  label = "paste(italic(median), \" = 0.10\")", parse = TRUE, color = "blue", size = 7, angle = 270)

ctnn1_tissues

#ggsave(filename = "output/Publication/Figures/Fig_3F.png",height = 10, width = , plot =ctnn1_tissues, dpi = 1200)
```



## Figure 3E


```{r}
file_dir = file.path("/Users/unawaz/Documents/2023/RNA-protein-correlationx/Data/")

data = list()

## Metadata
data$sample_md = read.delim("~/Documents/2023/RNA-protein-correlationx/Data/e0022_diann_051021_sample_mapping_averaged.txt")
data$sample_md_rep = read.table(file.path(file_dir, "e0022_diann_051021_sample_mapping_replicates.txt"), header=TRUE, sep = "\t")


data$prt_mtx =  read.delim(file.path(file_dir, "e0022_diann_051021_working_matrix_averaged.txt"),
                           header = TRUE, fill = TRUE, check.names = FALSE)

data$rna_mtx = read.delim(gzfile(file.path(file_dir, "rnaseq_voom.csv.gz")), header = TRUE,
                          fill = TRUE, sep = ",",
                          row.names = 1, check.names = FALSE)
```


```{r fig.height=6, fig.width=6, fig.cap= "Distribution of CTNNB1 mRNA expression across samples from the ProCan-DepMapSanger dataset, with outliers (> 3 standard deviations above the mean) indicated"}
library(ggbeeswarm)

rna_swarm_ctnnb1 =data$rna_mtx %>%
  t() %>%
  as.data.frame() %>%
  dplyr::select("CTNNB1") %>%
  scale() %>%
  as.data.frame() %>%
  mutate(Outlier = ifelse(CTNNB1 > 3, "Yes", "No")) %>%
  rownames_to_column("SIDM") %>%
  left_join(data$rna_mtx %>%
  t() %>%
  as.data.frame() %>%
  dplyr::select("RNA" = "CTNNB1") %>%
  rownames_to_column("SIDM"), by = "SIDM") %>%
  ggplot(aes(y = RNA, x='', color = Outlier)) +
  geom_beeswarm(cex=3, size=3) +
   scale_color_manual(values = c("No" = "#8d99ae",
                                "Yes" = "#bc4749"))  +
  theme(legend.position = "top") + theme_bw() +
    theme(axis.text.x = element_text(size = 20, family = "serif", color = "black"),
            axis.text.y = element_text(size = 20, family = "serif", color = "black"),
        axis.title = element_text(size=20, family = "serif"),
         legend.box.background = element_rect(color = "black"),
          legend.text = element_text(family = "serif", size=20),
          legend.title = element_text( family = "serif", size =20),
           plot.title = element_text(family = "serif", size =20),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line( size=.1 ),
              legend.position = "top",
        strip.text.y = element_text(
        size = 10, face = "bold.italic", family = "serif"
        ),
         strip.text.x = element_text(
        size = 10, family = "serif"
        )) + ggtitle("CTNNB1") + xlab("Samples") + ylab("RNA expression") + labs(color = "Outlier in RNA-seq")


rna_swarm_ctnnb1
ggsave(filename = "output/Publication/Figures/Fig_3E_RNA.png",height = 6, width = 5, plot = rna_swarm_ctnnb1, dpi = 1200 )

```


## Figure 3E

```{r}
ctnnb1= data$rna_mtx %>%
  t() %>%
  as.data.frame() %>%
  dplyr::select("CTNNB1") %>%
  scale() %>%
  as.data.frame() %>%
  mutate(Outlier = ifelse(abs(CTNNB1) > 3, "Yes", "No")) %>%
  rownames_to_column("SIDM") %>%
  left_join(data$rna_mtx %>%
  t() %>%
  as.data.frame() %>%
  dplyr::select("RNA" = "CTNNB1") %>%
  rownames_to_column("SIDM"), by = "SIDM") %>%
  left_join(data$prt_mtx %>%
  mutate(Project_Identifier = gsub(";.*", "", Project_Identifier)) %>%
  column_to_rownames("Project_Identifier") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("Protein") %>%
  mutate("Protein_name" = gsub(";.*", "", Protein)) %>%
  mutate("Gene" = mapIds(org.Hs.eg.db, keys = Protein_name,
                         keytype = "UNIPROT", column = "SYMBOL", multiVals="first")) %>%
  dplyr::select(-c(Protein_name, Protein)) %>%
  dplyr::filter(Gene == "CTNNB1") %>%
  column_to_rownames("Gene") %>%
  melt() %>% set_colnames(c("SIDM", "Protein"))) %>%
  ggplot(aes(x=RNA, y =Protein, color = Outlier)) + geom_point(size = 4) + theme_bw() + xlim(6,13) +
   scale_color_manual(values = c("No" = "#8d99ae",
                                "Yes" = "#bc4749")) +
    theme(axis.text.x = element_text(size = 20, family = "serif", color = "black"),
            axis.text.y = element_text(size = 20, family = "serif", color = "black"),
        axis.title = element_text(size=20, family = "serif"),
         legend.box.background = element_rect(color = "black"),
          legend.text = element_text(family = "serif"),
          legend.title = element_text( family = "serif"),
           plot.title = element_text(family = "serif", size =20),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line( size=.1 ),
              legend.position = "none",
        strip.text.y = element_text(
        size = 10, face = "bold.italic", family = "serif"
        ),
         strip.text.x = element_text(
        size = 10, family = "serif"
        )) + ggtitle("CTNNB1") + labs(x="RNA expression", y = "Protein abundance")

ggsave(filename = "output/Publication/Figures/Fig_3E_upper.png",height = 5.5, width = 5, plot = ctnnb1, dpi = 1200 )


 protein_swarm = data$rna_mtx %>%
  t() %>%
  as.data.frame() %>%
  dplyr::select("CTNNB1") %>%
  scale() %>%
  as.data.frame() %>%
  mutate(Outlier = ifelse(abs(CTNNB1) > 3, "Yes", "No")) %>%
  rownames_to_column("SIDM") %>%
  left_join(data$rna_mtx %>%
  t() %>%
  as.data.frame() %>%
  dplyr::select("RNA" = "CTNNB1") %>%
  rownames_to_column("SIDM"), by = "SIDM") %>%
  left_join(data$prt_mtx %>%
  mutate(Project_Identifier = gsub(";.*", "", Project_Identifier)) %>%
  column_to_rownames("Project_Identifier") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("Protein") %>%
  mutate("Protein_name" = gsub(";.*", "", Protein)) %>%
  mutate("Gene" = mapIds(org.Hs.eg.db, keys = Protein_name,
                         keytype = "UNIPROT", column = "SYMBOL", multiVals="first")) %>%
  dplyr::select(-c(Protein_name, Protein)) %>%
  dplyr::filter(Gene == "CTNNB1") %>%
  column_to_rownames("Gene") %>%
  melt() %>% set_colnames(c("SIDM", "Protein"))) %>%
  ggplot(aes(y = Protein, x='', color = Outlier)) +
  geom_beeswarm(size = 3, cex=3) +
scale_color_manual(values = c("No" = "#8d99ae",
                                "Yes" = "#bc4749"))  + theme_bw() +
    theme(axis.text.x = element_text(size = 20, family = "serif", color = "black"),
            axis.text.y = element_text(size = 20, family = "serif", color = "black"),
        axis.title = element_text(size=20, family = "serif"),
         legend.box.background = element_rect(color = "black"),
          legend.text = element_text(family = "serif"),
          legend.title = element_text( family = "serif"),
           plot.title = element_text(family = "serif", size =20),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line( size=.1 ),
              legend.position = "none",
        strip.text.y = element_text(
        size = 10, face = "bold.italic", family = "serif"
        ),
         strip.text.x = element_text(
        size = 10, family = "serif"
        )) + ylab("Protein abundance") + xlab("Samples")


ggsave(filename = "output/Publication/Figures/Fig_3E_lower.png",height = 5.5, width = 5, plot =protein_swarm, dpi = 1200 )
```


```{r fig.height=8, fig.width=6, fig.cap="Scatterplot showing mRNA expression and protein abundance for CTNNB1 (upper) and distribution of CTNNB1 protein abundance (lower) across samples in the ProCan-DepMapSanger dataset."}
ggarrange(ctnnb1, protein_swarm, nrow = 2)
```







